---
title: "BasicsOfFiveYearReview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BasicsOfFiveYearReview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an example vignette for the EFHSDM package. It provides an simplified example that includes several types of SDMs and an example of how to construct a weighted ensemble average. 
<!-- Note, revise these lines once the data is loading correctly -->

```{r setup}
library(EFHSDM)
library(akgfmaps)
```

# Load the data and add more info

```{r eval=F}
region.data <- region_data_all

bathy <- GOA_bathy
btemp <- GOA_btemp
btemp <- raster::crop(x = btemp, y = bathy)
slope <- GOA_slope
sponge <- GOA_sponge

region.data$j_rebs <- region.data$j_rebs + region.data$j_rough + region.data$j_bspot
region.data$a_rebs <- region.data$a_rebs + region.data$a_rough + region.data$a_bspot

region.data$sponge <- as.integer(region.data$sponge > 0)
region.data$coral <- as.integer(region.data$coral > 0)
region.data$pen <- as.integer(region.data$pen > 0)

region.data$logarea <- log(region.data$area)

lat <- raster::init(GOA_bathy, v = "y")
lat <- raster::mask(lat, GOA_bathy, overwrite = F)
lon <- raster::init(GOA_bathy, v = "x")
lon <- raster::mask(lon, GOA_bathy, overwrite = F)

raster.stack <- raster::stack(lon, lat, bathy, btemp, slope, sponge)
names(raster.stack) <- c("lon", "lat", "bdepth", "btemp", "slope", "sponge")

nice.names <- data.frame(
  stringsAsFactors = F,
  var = c(
    "lon", "lat", "bdepth", "slope", "aspectE", "aspectN", "curve", "btemp", "speed", "tmax", "BPI", "phi", "vitalrate",
    "color", "sponge", "coral", "pen", "area", "bcurrentU", "bcurrentV", "bcurrentUSD", "bcurrentVSD", "lon*lat",
    "bcurrentU*bcurrentV", "bcurrentUSD*bcurrentVSD", "aspectE*aspectN", "rocky"
  ),
  name = c(
    "Longitude", "Latitude", "Depth (m)", "Slope (degrees)", "Slope Eastness", "Slope Northness",
    "Terrain Curvature", "Temperature (C)", "Current speed (m/s)",
    "Tidal Maximum (cm/s)", "BPI", "Sediment Grain size (phi)", "Growth Potential (g/g/day)",
    "Ocean color", "Sponge presence", "Coral presence", "Pennatulacean presence", "Area Swept",
    "Current Velocity East (m/s)", "Current Velocity North (m/s)", "Current Velocity East SD (m/s)",
    "Current Velocity North SD (m/s)", "Position (lon,lat)", "Current Velocity (east,north)",
    "Current Velocity SD", "Slope Aspect (east,north)", "Bottom Rockiness (%)"
  )
)

species <- "a_atf"
figure.name <- "Adult arrowtooth flounder"
```

For the example, we will restrict the dataset to the last ten years, note that some species were not adequately identified until a particular year and the dataset should be subsetted accordingly.

```{r eval=F}
start.year <- 2012

species.data <- subset(region.data, year >= start.year)

# Assign the CV folds from here; the EFH project used the RACE code to assign seed
set.seed(10110)

random.folds <- rep(LETTERS[1:10], length.out = nrow(species.data))
species.data$Folds <- sample(random.folds, size = nrow(species.data), replace = F)

# a good start is to map the presence data
hd <- quantile(species.data[species.data[, species] > 0, species], .9)

MakeAKGFDotplot(
  presence = species.data[species.data[, species] > 0, ],
  absence = species.data[species.data[, species] == 0, ],
  highdensity = species.data[species.data[, species] >= hd, ],
  dataCRS = raster.stack@crs, region = "goa", title.name = figure.name
)

gam.form <- formula("a_atf ~ s(lon,lat,bs = 'ds',m=c(1,.5), k=10) + s(bdepth, bs='tp',m=1,k=4) +
                    s(btemp, bs='tp',m=1,k=4) + s(slope, bs='tp',m=1,k=4) + as.factor(sponge) + offset(logarea)")
```


# Start running the MaxEnt model

 In practice, one should try several different values for `regmult`, and test the best one. For now we'll just use `1`. Also note that `maxnet` models frequently fail to converge.

```{r eval=F}
maxnet.model <- FitMaxnet(
  data = species.data, species = species,
  vars = c("lon", "lat", "btemp", "bdepth", "slope"), facs = "sponge",
  regmult = 1, reduce = T
)

# the results of a maxnet model should always be scaled, if one wishes to make abundance predictions
maxnet.scale <- mean(species.data[, species]) / mean(exp(predict(maxnet.model, newdata = species.data, type = "link") +
  maxnet.model$ent))

maxnet.abund <- MakeMaxEntAbundance(
  model = maxnet.model, maxent.stack = raster.stack,
  scale.fac = maxnet.scale, type = "cloglog"
)

# Now crossvalidate and get fit metrics
maxnet.cv <- CrossValidateModel(
  model = maxnet.model, data = species.data, species = species, group = "Folds",
  model.type = "maxnet", key = "hauljoin", scale.preds = T, regmult = 1
)
maxnet.errors <- maxnet.cv[[1]]
maxnet.cv.models <- maxnet.cv[[2]]

maxnet.rmse <- RMSE(pred = maxnet.errors$cvpred, obs = maxnet.errors$abund)

# get the effects and confidence intervals
maxnet.effects <- GetMaxnetEffects(
  model = maxnet.model, cv.models = maxnet.cv.models,
  vars = "all", add.entropy = T, scale = "log", data = species.data,
  maxnet2d = list(c("lon", "lat")), scale.factor = maxnet.scale
)
```

We could go through and get deviance and variance estimates for the model now, but it's better to wait and see which models actually are included in the ensemble, in order to save on processing time.

# Cloglong paGAM

```{r eval=F}
cloglog.model <- FitGAM(
  gam.formula = gam.form, data = species.data,
  family.gam = "binomial", link.fx = "cloglog", reduce = T, select = T,
  verbose = F
)

# this type of model should also always be scaled if abundance prediction is desired
cloglog.scale <- mean(species.data[, species]) / mean(exp(predict(cloglog.model, type = "link")))

cloglog.abund <- MakeGAMAbundance(
  model = cloglog.model, r.stack = raster.stack, scale.factor = cloglog.scale,
  filename = ""
)

cloglog.cv <- CrossValidateModel(
  model = cloglog.model, model.type = "cloglog", data = species.data, scale.preds = T,
  species = species, group = "Folds", key = "hauljoin"
)
cloglog.errors <- cloglog.cv[[1]]
cloglog.cv.models <- cloglog.cv[[2]]

cloglog.effects <- GetGAMEffects(
  model = cloglog.model, data = species.data, vars = "all", scale = "log",
  cv.model.list = cloglog.cv.models, scale.factor = cloglog.scale
)

cloglog.rmse <- RMSE(pred = cloglog.errors$cvpred, obs = cloglog.errors$abund)
```

#Hurdle GAM

ziplss GAMs require a separate formula for the probability part of the model, and it should leave off the dependent variable
```{r eval=F}
prob.form <- formula("~ s(lon,lat,bs = 'ds',m=c(1,.5), k=10) + s(bdepth, bs='tp',m=1,k=4) +
                    s(btemp, bs='tp',m=1,k=4) + s(slope, bs='tp',m=1,k=4) + as.factor(sponge) + offset(logarea)")

hpoisson.model <- FitHurdleGAM(
  density.formula = gam.form, prob.formula = prob.form,
  data = species.data, verbose = F, select = T, reduce = T
)

# scaling a hurdle model is optional, but we choose to do it for consistency
hpoisson.scale <- mean(species.data[, species]) / mean(predict(hpoisson.model, type = "response"))
hpoisson.abund <- MakeGAMAbundance(
  model = hpoisson.model, r.stack = raster.stack, scale.factor = hpoisson.scale,
  filename = ""
)

hpoisson.cv <- CrossValidateModel(
  model = hpoisson.model, model.type = "hgam", data = species.data,
  species = species, group = "Folds", key = "hauljoin", scale.preds = T
)

hpoisson.errors <- hpoisson.cv[[1]]
hpoisson.cv.models <- hpoisson.cv[[2]]

hpoisson.effects <- GetGAMEffects(
  model = hpoisson.model, data = species.data, cv.model.list = hpoisson.cv.models,
  scale.factor = hpoisson.scale, vars = "all", scale = "log"
)

hpoisson.rmse <- RMSE(pred = hpoisson.errors$cvpred, obs = hpoisson.errors$abund)
```

# Poisson GAM

```{r eval=F}
poisson.model <- FitGAM(
  gam.formula = gam.form, data = species.data, verbose = F,
  reduce = T, select = T, family.gam = "poisson"
)

# by definition, the mean of a poisson model will always be equal to the mean of the data, so the scale is always 1
poisson.scale <- mean(species.data[, species]) / mean(predict(poisson.model, type = "response"))

poisson.abund <- MakeGAMAbundance(
  model = poisson.model, r.stack = raster.stack, scale.factor = poisson.scale,
  filename = ""
)

poisson.cv <- CrossValidateModel(
  model = poisson.model, model.type = "gam", species = species, data = species.data,
  group = "Folds", key = "hauljoin", scale.preds = T
)
poisson.errors <- poisson.cv[[1]]
poisson.cv.models <- poisson.cv[[2]]


poisson.effects <- GetGAMEffects(
  model = poisson.model, data = species.data, vars = "all", scale = "log",
  cv.model.list = poisson.cv.models, scale.factor = poisson.scale
)

poisson.rmse <- RMSE(pred = poisson.errors$cvpred, obs = poisson.errors$abund)
```

# Negative binomial GAM

```{r eval=F}
negbin.model <- FitGAM(
  gam.formula = gam.form, data = species.data, verbose = F, select = T,
  reduce = T, family.gam = "nb"
)

# again, scaling the negbin model is optional, but
negbin.scale <- mean(species.data[, species]) / mean(predict(negbin.model, type = "response"))
negbin.abund <- MakeGAMAbundance(
  model = negbin.model, r.stack = raster.stack, scale.factor = negbin.scale,
  filename = ""
)

negbin.cv <- CrossValidateModel(
  model = negbin.model, model.type = "gam", species = species, data = species.data,
  group = "Folds", key = "hauljoin", scale.preds = T
)

negbin.errors <- negbin.cv[[1]]
negbin.cv.models <- negbin.cv[[2]]


negbin.effects <- GetGAMEffects(
  model = negbin.model, data = species.data, vars = "all", scale = "log",
  cv.model.list = negbin.cv.models, scale.factor = negbin.scale
)

negbin.rmse <- RMSE(pred = negbin.errors$cvpred, obs = negbin.errors$abund)
```

# Model ensemble

All of the models converged and passed the abundance check, so we will test all of them in the ensemble.

```{r eval=F}
# assemble some things
rmse.vec <- c(maxnet = maxnet.rmse, cloglog = cloglog.rmse, hpoisson = hpoisson.rmse, poisson = poisson.rmse, negbin = negbin.rmse)

# by default, the ensemble will include only one of each type
model.types <- c("maxnet", "cloglog", "hgam", "gam", "gam")

model.weights <- MakeEnsemble(rmse = rmse.vec, minimum = .1, model.types = model.types)

##############################
# so now we know there are four models, we can go back and get the relevant features
# Like the variance maps (Note, this step takes awhile ~ 45 minutes; you might want to skip it)
maxnet.variance <- MakeVarianceRasters(
  model.list = maxnet.cv.models, raster.stack = raster.stack,
  model.type = "maxnet", scale.factor = maxnet.scale
)
cloglog.variance <- MakeVarianceRasters(
  model.list = cloglog.cv.models, raster.stack = raster.stack,
  model.type = "cloglog", scale.factor = cloglog.scale
)
hpoisson.variance <- MakeVarianceRasters(
  model.list = hpoisson.cv.models, raster.stack = raster.stack,
  model.type = "hgam", scale.factor = hpoisson.scale
)
poisson.variance <- MakeVarianceRasters(
  model.list = poisson.cv.models, raster.stack = raster.stack,
  model.type = "gam", scale.factor = poisson.scale
)

##############################
# and the deviance estimates
# in order to keep the deviance estimates consistent with the GAM, specify a "maxnet2d" argument as a list
maxnet.devs <- MaxnetStats(
  model = maxnet.model, data = species.data, species = species, regmult = 1,
  maxnet2d = list(c("lon", "lat"))
)
maxnet.coefs <- MaxnetCoefs(maxnet.model, maxnet2d = list(c("lon", "lat")))

cloglog.devs <- GAMStats(model = cloglog.model, data = species.data)
hpoisson.devs <- GAMStats(model = hpoisson.model, data = species.data)
poisson.devs <- GAMStats(model = poisson.model, data = species.data)

# take a weighted average for the ensemble deviance explained
dev.dat <- rbind(
  maxnet.devs,
  c(cloglog.devs, slope = 0), # slope was dropped from the cloglog model, so it explains 0
  hpoisson.devs,
  poisson.devs
)
apply(dev.dat, MARGIN = 2, FUN = "weighted.mean", w = model.weights[model.weights > 0], na.rm = T)

##############################
# now make the actual ensemble map
# the list of abundances ought to include all non-zero weight models
abund.list <- list(maxnet = maxnet.abund, cloglog = cloglog.abund, hpoisson = hpoisson.abund, poisson = poisson.abund)

ensemble.abund <- MakeEnsembleAbundance(model.weights = model.weights, abund.list = abund.list)

MakeAKGFDensityplot(
  region = "goa", density.map = ensemble.abund, buffer = .98, legend.title = "Abundance",
  title.name = figure.name
)

# this finds the ensemble efh breaks
# the threshold .0513 is equivalent to 5% encounter probability under a poisson distribution
ensemble.breaks <- FindEFHbreaks(
  abund.raster = ensemble.abund, method = "percentile", threshold = .0513,
  data = species.data, quantiles = c(.05, .25, .5, .75)
)

##############################
# and the efh map
ensemble.efh <- raster::cut(ensemble.abund, breaks = ensemble.breaks)

MakeAKGFEFHplot(region = "goa", efh.map = ensemble.efh, title.name = figure.name)

# this gives the area of the 95% EFH in km2
sum(raster::getValues(ensemble.efh) > 1, na.rm = T)

############################
# this calculates the ensemble effects from the effects lists produced earlier
effects.list <- list(maxnet.effects, cloglog.effects, hpoisson.effects, poisson.effects)
ensemble.effects <- GetEnsembleEffects(
  effects.list = effects.list, model.weights = model.weights,
  vars = "all", scale = "log"
)

ensemble.plots <- Effectsplot(
  effects.list = ensemble.effects, region = "goa", crs = raster.stack@crs,
  nice.names = nice.names, vars = "all"
)
gridExtra::grid.arrange(grobs = ensemble.plots, ncol = 3, nrow = 2)


# this makes a useful data frame of the ensemble predictions
errors.list <- list(maxnet = maxnet.errors, cloglog = cloglog.errors, hpoisson = hpoisson.errors, poisson = poisson.errors)

ensemble.preds <- ValidateEnsemble(
  pred.list = errors.list, model.weights = model.weights, output = T, make.plots = T,
  key = "hauljoin", latlon = T, group = "Folds", method = "spearman"
)

##############################
# ensemble variance & cv map
var.list <- list(maxnet = maxnet.variance, cloglog = cloglog.variance, hpoisson = hpoisson.variance, poisson = poisson.variance)
ensemble.var <- GetEnsembleVariance(
  model.weights = model.weights, variance.list = var.list, abund.list = abund.list,
  ensemble.abund = ensemble.abund
)

ensemble.cv <- sqrt(ensemble.var) / ensemble.abund
MakeAKGFDensityplot(
  region = "goa", density.map = ensemble.cv, buffer = .98, legend.title = "CV of abundance",
  title.name = figure.name
)

# All Done. This covers the essentials produced during the 2022 EFH 5-year Review.
```
